### 2022-11-09 NSF2022/Description/specificaim2.tex

%\label{subsec:specific_aim_2}
%\begin{wrapfigure}[14]{r}{0.61\textwidth}
%  \centering{\includegraphics[width=0.60\textwidth]{Figures/domain.pdf}}
%    \vspace{-15pt}
%    \caption{\label{fig:flow_map} \footnotesize A collection of 3D
%    amphiphilic particles suspended in a solvent. The solvent velocity
%    satisfies the Stokes equations, and the hydrophobic forces depend on
%    the solvent activity that satisfies the screened Laplace equation.
%    The geometry is updated by solving the mobility problem.} 
%\end{wrapfigure}
The HAP model requires solving exterior problems for the order
parameter~\eqref{eqn:phase} and the Stokes equations~\eqref{eqn:stokes}.
We will develop methods based on combinations of volume and layer
potentials, and inexact Newton's method. As mentioned in
Section~\ref{sec:bie}, numerical methods to solve these integral
equations (IEs) have several advantages over their PDE-based
counterpart. The PIs have demonstrated that IEs are a powerful tool to
simulate two-dimensional suspensions of amphiphilic
particles~\cite{Fu2018_SIAM, FuQuRyYo22, fu-ryh-qua-you2022}. Moreover,
elements of these formulations have been described in three
dimensions~\cite{ying_2006, manasthesis, rac-gre2016}. This proposal
will extend the two-dimensional methods to more general amphiphilic
particle models and also to three dimensions. In addition, the methods
will allow us to use results we have shown for $\mathcal{O}(100)$
grains, and extend them to suspensions with 100X--1000X more grains.
These methods will be motivated by models and analysis developed in
Specific Aims 1 and 3. \S\ref{subsec:AC} describes how the non-linear
Allen-Cahn equation~\eqref{eq:SL} will be iteratively solved using an IE
formulation. Then, \S\ref{subsec:NumericalIssues} describes how we will
combine accurate quadrature and fast summation methods to accelerate the
time to accurate solutions. A natural extension to create simulations
that are realizable by experimentalists is to include confinement, and
this is discussed in \S\ref{subsec:Confinement}. This extension will
also allow us to study new rheological properties of the suspensions.
Finally, \S\ref{subsec:timeStepping} proposes collision-aware adaptive
time stepping methods.

@@

The PIs have developed numerical methods to solve~\eqref{eqn:phase} when
$f(u) = u^2$---this choice results in the linear screened Laplace
equation. For a more general $f$, such as a double well potential, we
will interpret the solution of~\eqref{eqn:phase} as the root finding
problem for
\begin{align}
  \label{eq:F}
  J[u] = -\rho^2 \Delta u + \tfrac{1}{2}f'(u).
\end{align}
This non-linear PDE will be solved with Newton's method
\begin{align*}
  u^{N+1} = u^{N} - \alpha (dJ^N)^{-1} J(u^N),
\end{align*}
where $dJ^N = -\rho^2 \Delta + \tfrac{1}{2}f''(u^N)$ is the Fr\'{e}chet
derivative of $F$, and $\alpha$ is the step size which will be
determined using backtracking. As expected, the biggest computational
challenge is solving $- \rho^2 \Delta v + \tfrac{1}{2}f''(u^{N}) v =
J[u^N]$. Solving this variable coefficient PDE with an integral equation
is challenging since a fundamental solution is not readily available.
However, since $f''(u_0) = f''(u_1) = 2a(u_1 - u)0)^2 + 2c$, f' We
propose to instead use an inexact Newton's method where $dJ^{-1}$ is
defined in terms of the continuous, piecewise quadratic approximation
\begin{align*}
  f(u) \approx \tilde{f}(u) = bu + \begin{cases}
    a(u - u_0)^2, &\mbox{if } u \leq (u_0 + u_1)/2, \\
    (u - u_1)^2 + \tfrac{1}{4}(a-1)(u_0-u_1)^2 , &\mbox{if } u > (u_0 + u_1)/2,
  \end{cases}
\end{align*}
where $a>0$, and the resulting inexact Newton's method requires solving
the piecewise constant PDE
\begin{align}
  \label{eq:screenedPoisson}
  -\rho^2 \Delta v + \tfrac{1}{2}\tilde{f}''(u^{N})v = J[u^N],
\end{align}
which is a screened Poisson equation with piecewise a constant screening
coefficient. We test the efficacy of exact and inexact Newton's method
applied to~\eqref{eq:F} in $(0,1)$ by discretizing $dJ$ with a
second-order finite difference method. The choice of parameters $a$,
$b$, $u_0$, and $u_1$ result in the double well potential illustrated in
Figure~\ref{fig:CA}(a), and both the exact and inexact Newton's methods
converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the max norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
piecewise constant approximate (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach of a tolerance
of $10^{-4}$, each iteration is significantly cheaper in higher
dimensions.

@@

% ----------------------------------------------------------------------
\subsubsection{Increased accuracy and reduced CPU time}
\label{subsec:NumericalIssues}
While IE formulations of PDEs have several advantages, several
challenges must be overcome before the code can be used to simulate
systems with large numbers of grains. For example, when grains are close
to one another, which is inevitable in dense suspensions, the resulting
integrals are nearly-singular. This issue is well-known in the IE
community and several algorithms have been developed in the last 30
years. An incomplete list of methods that address near-singular
integration include panel-based methods~\cite{helsing_2008a,
helsing_tutorial_2012}, Quadrature by Expansions
(QBX)~\cite{din-hua-mar-tan2021, klo-bar-gre-one2013, ste-bar2022,
af2018sisc}, and interpolation-based methods~\cite{ying_2006,
qua-bir2014, rah-vee-zor-bir2015}. The proposed research, which is
described next,  will combine several methods including singularity
subtraction, analytic function theory, and spherical basis
representations.

@@

In three dimensions, the quadrature methods are much more challenging.
Singularity subtraction methods can be applied, but analytic function
theory is no longer applicable. Singularity subtraction methods have
been applied using, for example, smoothed variations of
the~\cite{beale1, cor2001, tlu-bea2019}. Alternatively, efficient
quadrature methods can be applied directly to accurately compute
singular and near-singular integrals~\cite{khayat_2005,
gimbutas2013sisc}. However, since we are interested in round (2D) and
spherical (3D) grains, we will exploit the fact that integrals can be
computed exactly after transforming the integrals to to a Fourier (2D)
or spherical harmonic (3D) basis. This technique can be used to evaluate
layer potentials for both Stokes~\cite{cor-vee2018} and the screened
Laplace equation~\cite{koh-cor-che-vee2021}.

@@

% ----------------------------------------------------------------------
\subsubsection{Incorporating confinement}
\label{subsec:Confinement}
In experiments, it is known that interactions between bounding walls and
grains cannot be ignored~\cite{qua-gan-you2021,
KaouiHartingMisbah2011_PRE, her-sto-gra2005}. Therefore, the necessary
modifications to the code will be performed so that confinement can be
considered. PI BQ has considered suspensions with confinement for
several projects including dynamic suspensions~\cite{qua-bir2014,
bys-sha-qua2020} and porous media~\cite{chi-moo-qua2020,
dea-qua-bir-jua2018, moo-che-chi-qua2022, qua-gan-you2021}.

>>>>

%\label{subsec:specific_aim_2}
%\begin{wrapfigure}[14]{r}{0.61\textwidth}
%  \centering{\includegraphics[width=0.60\textwidth]{Figures/domain.pdf}}
%    \vspace{-15pt}
%    \caption{\label{fig:flow_map} \footnotesize A collection of 3D
%    amphiphilic particles suspended in a solvent. The solvent velocity
%    satisfies the Stokes equations, and the hydrophobic forces depend on
%    the solvent activity that satisfies the screened Laplace equation.
%    The geometry is updated by solving the mobility problem.} 
%\end{wrapfigure}
The HAP model requires solving exterior problems for the order
parameter~\eqref{eqn:phase} and the Stokes equations~\eqref{eqn:stokes}.
We will develop methods based on combinations of volume and layer
potentials, and inexact Newton's method. As mentioned in
Section~\ref{sec:bie}, numerical methods to solve these integral
equations (IEs) have several advantages over their PDE-based
counterpart. The PIs have demonstrated that IEs are a powerful tool to
simulate two-dimensional suspensions of amphiphilic
particles~\cite{Fu2018_SIAM, FuQuRyYo22, fu-ryh-qua-you2022}. Moreover,
elements of these formulations have been described in three
dimensions~\cite{ying_2006, manasthesis, rac-gre2016}. This proposal
will extend the two-dimensional methods to more general amphiphilic
particle models and also to three dimensions. In addition, the methods
will allow us to use results we have shown for $\mathcal{O}(100)$
granules, and extend them to suspensions with 100X--1000X more granules.
These methods will be motivated by models and analysis developed in
Specific Aims 1 and 3. \S\ref{subsec:AC} describes how the non-linear
equation for the order parameter~\eqref{eqn:phase} will be iteratively
solved using an IE formulation. Then, \S\ref{subsec:NumericalIssues}
describes how we will combine accurate quadrature and fast summation
methods to accelerate the time to accurate solutions. A natural
extension to create simulations that are realizable by experimentalists
is to include confinement, and this is discussed in
\S\ref{subsec:Confinement}. This extension will also allow us to study
new rheological properties of the suspensions.  Finally,
\S\ref{subsec:timeStepping} proposes collision-aware adaptive time
stepping methods.

@@

The PIs have developed numerical methods to solve~\eqref{eqn:phase} when
$f(u) = u^2$---this choice results in the linear screened Laplace
equation. For a more general double well potential $f$, with local
minimums at $u_0$ and $u_1$, we will interpret the solution
of~\eqref{eqn:phase} as the root finding problem for
\begin{align}
  \label{eqn:phaseRoot}
  J[u] = -\rho^2 \Delta u + \tfrac{1}{2}f'(u).
\end{align}
This non-linear PDE will be solved with Newton's method, $u^{N+1} =
u^{N} - \alpha (dJ^N)^{-1} J(u^N)$, where $dJ^N = -\rho^2 \Delta +
\tfrac{1}{2}f''(u^N)$ is the Fr\'{e}chet derivative of $F$, and $\alpha$
is the step size which will be determined using backtracking. As
expected, the biggest computational challenge is solving $-\rho^2 \Delta
v + \tfrac{1}{2}f''(u^{N}) v = J[u^N]$. Solving this variable
coefficient PDE with an integral equation is challenging since a
fundamental solution is not readily available. However, by choosing a
double well potential with $f''(u_0) = f''(u_1) = \beta$, we propose to
instead use an inexact Newton's method where $dJ^{-1}$ is approximated
by solving $-\rho^2 \Delta v + \tfrac{beta}{2} v = J[u^N]$. We test the
efficacy of exact and inexact Newton's method applied
to~\eqref{eqn:phaseRoot} in $(0,1)$ by discretizing $dJ$ with a
second-order finite difference method. The double well potential is
illustrated in Figure~\ref{fig:CA}(a), and both the exact and inexact
Newton's methods converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the max norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
approximate second derivative (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach of a tolerance
of $10^{-4}$, each iteration is significantly cheaper to apply.
\todo[inline]{put pics here}

@@

% ----------------------------------------------------------------------
\subsubsection{Increased accuracy and reduced CPU time}
\label{subsec:NumericalIssues}
While IE formulations of PDEs have several advantages, several
challenges must be overcome before the code can be used to simulate
systems with large numbers of granules. For example, when granules are
close to one another, which is inevitable in dense suspensions, the
resulting integrals are nearly-singular. This issue is well-known in the
IE community and several algorithms have been developed in the last 30
years. An incomplete list of methods that address near-singular
integration include panel-based methods~\cite{helsing_2008a,
helsing_tutorial_2012}, Quadrature by Expansions
(QBX)~\cite{din-hua-mar-tan2021, klo-bar-gre-one2013, ste-bar2022,
af2018sisc}, and interpolation-based methods~\cite{ying_2006,
qua-bir2014, rah-vee-zor-bir2015}. The proposed research, which is
described next,  will combine several methods including singularity
subtraction, analytic function theory, and spherical basis
representations.

@@

In three dimensions, the quadrature methods are much more challenging.
Singularity subtraction methods can be applied, but analytic function
theory is no longer applicable. Singularity subtraction methods have
been applied using, for example, smoothed variations of
the~\cite{beale1, cor2001, tlu-bea2019}. Alternatively, efficient
quadrature methods can be applied directly to accurately compute
singular and near-singular integrals~\cite{khayat_2005,
gimbutas2013sisc}. However, since we are interested in round (2D) and
spherical (3D) granules, we will exploit the fact that integrals can be
computed exactly after transforming the integrals to to a Fourier (2D)
or spherical harmonic (3D) basis. This technique can be used to evaluate
layer potentials for both Stokes~\cite{cor-vee2018} and the screened
Laplace equation~\cite{koh-cor-che-vee2021}.

@@

% ----------------------------------------------------------------------
\subsubsection{Incorporating confinement}
\label{subsec:Confinement}
In experiments, it is known that interactions between bounding walls and
granules cannot be ignored~\cite{qua-gan-you2021,
KaouiHartingMisbah2011_PRE, her-sto-gra2005}. Therefore, the necessary
modifications to the code will be performed so that confinement can be
considered. PI BQ has considered suspensions with confinement for
several projects including dynamic suspensions~\cite{qua-bir2014,
bys-sha-qua2020} and porous media~\cite{chi-moo-qua2020,
dea-qua-bir-jua2018, moo-che-chi-qua2022, qua-gan-you2021}.
